Datastorr: a workflow and package for delivering successive versions of ``living data'' directly into R

### 2018-10-03 ms.tex
minor edits throughout

\begin{frontmatter}
\maketitle
\begin{abstract}
The sharing and re-use of data has become a cornerstone of modern science. Multiple platforms now allow quick and easy data sharing. So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created. In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing Users to retrieve and load different versions of a dataset directly into the R platform. Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHuB and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets. Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost and with maximal performance. 
\end{abstract}

@@

An evolving (or `living') dataset is one that is subject to recurrent or occasional change.  Typical changes may include improving the quality of existing data, adding new data, re-structuring the dataset content, or integrating with other datasets. For example, a dataset on biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. Moreover, change may not be a temporary state, rather the dataset continue to evolve over extended periods \cite[e.g.][]{Ernest-2018}. Evolving datasets are never ``finished'', and as such there is no ``master'' or ``canonical'' version. Rather, as research around a data product grows, there might be many valid versions produced.  Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.

A common approach taken by those maintaining an evolving dataset is to release sequential versions of the dataset, each containing a snapshot of the dataset at the time of release \cite[e.g.][]{Falster-2015, Pennell-2015a, Yenni-2018, Abolfathi-2018}. Ideally, the latest versions of a dataset would be immediately available to global users, along with notes describing the changes when compared to previous versions. For the sake of reproducibility, previous versions of the dataset should also remain available. Historically, small research groups might have solved the issue of versioning data internally, for example by emailing around the latest version. However, as science moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.

@@

In brief, the workflow we present here borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset. Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig. \ref{fig:technology_stack}). Each of these will likely have different goals and requirements (see Table \ref{tab:user_requirements}). When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer. Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e. loaded) on a User's local computer. In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.

@@

Version control, primarily an open-source variety called \texttt{git}, has become widespread in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and also data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}. \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g. ``.csv'', ``.tsv'', ``.txt''; with some tricks git can also indicate changes in some other file types such as ``.xlsx''). It  allows users to annotate commits with informative messages detailing the rationale for those changes.  The ``history'' of commits is also visible to anyone interacting with the repository. In its present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.

@@

A central feature of the proposed system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}. Multiple Developers can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. The proposed system of data delivery thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

@@

In the long term, scientists want their datasets, software, and papers to preserved and remain accessible. While our proposed system for data delivery does not guarantee long-term preservation, users can also choose to automatically archive data-versions released on \texttt{GitHub} version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release. Currently, both \texttt{Zenodo} and  \texttt{FigShare}each integrate with \texttt{GitHub} for archiving of material hosted there. Ideally, tools like \texttt{datastorr} would also be developed to pull versions from these archives too.

>>>>

\begin{frontmatter}
\maketitle
\begin{abstract}
The sharing and re-use of data has become a cornerstone of modern science. Multiple platforms now allow quick and easy data sharing. So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created. In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing users to retrieve and load different versions of a dataset directly into the R platform. Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHub and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets. Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost. 
\end{abstract}

@@

An evolving (or `living') dataset is one that is subject to recurrent or occasional change.  Typical changes may include improving the quality of existing data, adding new data, re-structuring the dataset content, or integrating with other datasets. For example, a dataset on biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. In some cases, datasets may be expected to continue to evolve over extended periods \cite[e.g.][]{Ernest-2018}. Evolving datasets are never ``finished'', and as such there is no ``master'' or ``canonical'' version. Rather, as research around a data product grows, there might be many valid versions produced.  %Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.

A common approach taken by those maintaining an evolving dataset is to release sequential versions of the dataset, each containing a snapshot of the dataset at the time of release \cite[e.g.][]{Falster-2015, Pennell-2015a, Yenni-2018, Abolfathi-2018}. Ideally, the latest versions of a dataset would be immediately available to all users across the globe, along with notes describing the changes when compared to previous versions. For the sake of reproducibility, previous versions of the dataset should also remain available. Historically, small research groups might have solved the issue of versioning data internally, for example by emailing around the latest version. However, as science moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.

@@

In brief, the workflow we present here borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset. Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig. \ref{fig:technology_stack}). Each of these will likely have different goals and requirements (see Table \ref{tab:user_requirements}). When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer. Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a User's local computer. In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.

@@

Version control, primarily an open-source variety called \texttt{git}, has become widespread in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and also data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}. \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g., ``.csv'', ``.tsv'', ``.txt''; with some tricks git can also indicate changes in some other file types such as ``.xlsx''). It  allows users to annotate commits with informative messages detailing the rationale for those changes.  The ``history'' of commits is also visible to anyone interacting with the repository. In its present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.

@@

A central feature of the proposed system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}. Multiple Developers can make changes to different parts of the code (or, in our case, data) and the \texttt{git} system will integrate these together or, when needed, flag where there are conflicts that need to be resolved. The proposed system of data delivery thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

@@

In the long term, scientists want their datasets, software, and papers to preserved and remain accessible. While our proposed system for data delivery does not guarantee long-term preservation, users can also choose to automatically archive data-versions released on \texttt{GitHub} version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release. Currently, both \texttt{Zenodo} and  \texttt{FigShare} each integrate with \texttt{GitHub} for archiving of material hosted there. Ideally, tools like \texttt{datastorr} would also be developed to pull versions from these archives too.
